---
title: "opttext"
output: github_document
---

**opttext** is an R package for high-performance text preprocessing, document-term matrix construction, and exploratory text analysis. The package is designed for workflows that need efficient handling of large text corpora while minimizing memory overhead and unnecessary intermediate objects.

## Motivation and problem scope

Many text analysis workflows rely on token-based pipelines that create intermediate objects before constructing document-term matrices. While flexible, these approaches can become computationally expensive when working with large corpora.

`opttext` focuses on common preprocessing and matrix-construction tasks by operating directly on character vectors whenever possible and by using sparse matrix representations throughout the workflow.

## Features

The package currently provides:

 1. Fast text preprocessing functions.
 2. High-performance document-term matrix construction.
 3. Sparse matrix trimming for feature reduction.
 4. Exploratory visualizations for text data.
 5. Example datasets for testing and demonstration.

## Installation

```{r eval=FALSE}
devtools::install_github("VikyRackova/opttext")
```

## Example workflow

The typical workflow consists of preprocessing text, constructing a sparse document-term matrix, removing infrequent terms, and visualizing the resulting corpus.

```{r eval=FALSE}
texts <- c(
  "Inflation increased during this quarter.",
  "Economic growth remained stable.",
  "Inflation growth belongs to important indicators."
)

clean_texts <- lowercase(texts)
dtm <- fast_dtm(clean_texts)
dtm_small <- dtm_trim(dtm, min_docfreq = 2, min_termfreq = 2)

plot_top_terms(dtm_small)
```

The output of `fast_dtm()` is compatible with sparse-matrix-based workflows and can be used directly with packages such as:

```{r eval=FALSE}
topicmodels::LDA(dtm, k = 10)
```

## Included datasets

The package includes curated text datasets that serve both as examples and as practical test cases for evaluating preprocessing performance, document-term matrix construction, and exploratory text-analysis techniques.

#### FED_Minutes

Federal Open Market Committee meeting minutes from March 1994 to November 2024.

```{r eval=FALSE}
data("FED_Minutes")
```

Variables:

-   Date
-   Text

The dataset can be used for testing preprocessing functions, benchmarking matrix construction, and demonstrating text-analysis workflows.

---

## Relationship to other packages

`opttext` is designed to complement existing text-analysis workflows rather than replace them. Its functions follow concepts familiar to users of packages such as `quanteda`, `tidytext`, and `tm`, while emphasizing computational efficiency and scalability.

For example, `lowercase()` performs the same type of text normalization commonly used in text-analysis pipelines but incorporates parallel processing to improve performance on larger corpora. Similarly, `fast_dtm()` follows the logic of traditional document-term matrix constructors while returning a sparse S4 object that can be used directly in downstream applications, including topic modeling methods such as latent Dirichlet allocation (LDA).

In benchmark tests using the FED_Minutes dataset, `fast_dtm()` substantially outperformed a comparable `quanteda::tokens() + quanteda::dfm()` workflow when constructing sparse document-term matrices directly from character vectors. Performance gains for preprocessing tasks depend on the structure and size of the input data. In the included benchmarks, `opttext::lowercase()` outperformed base R's `tolower()`, while `stringi::stri_trans_tolower()` remained faster on long real-world documents. `remove_punctuation()` outperformed comparable implementations based on `stringi`. `remove_numbers()` performed similarly to equivalent `stringi` workflows. The combined preprocessing pipeline generally outperformed comparable workflows based on base R and `quanteda`. 

Benchmark script, `inst\benchmark.R`, is included in the repository to facilitate transparent comparisons with established alternatives.

## Documentation

All exported functions include documentation with examples, arguments, and return values. For usage details, see the package help pages after installation.

## Author

Viktoria Račková\
Master's Programme in Financial and Economic Research specialization Econometrics\
Maastricht University
